The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel {\normalsize \ldots after that it gets a little tricky.}

### 2013-03-16 trunk/tasbot/paper/paper.tex
More about weighting objectives. Some figures

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}

@@

\section*{Introduction}
The Nintendo Entertainment System is probably the best video game console, citation {\it not} needed. Like many, I have spent thousands of hours of my life playing NES games, including several complete playthroughs of classics like Super Mario Bros., Bionic Commando, Bubble Bobble, and other favorites. By the year 2013, home computers have become many orders of magnitude faster and more capacious than the NES hardware. This suggested to me that it may be time to automate the playing of NES games, in order to save time.\footnote{Rather, to replace it with time spent programming.} In this paper I present a generic technique for automating the playing of NES games. The approach is practical on a single computer, and succeeds on some games, such as Super Mario Bros.. The approach is amusingly elegant and surprisingly effective, requires no detailed knowledge of the game being played, and is capable of novel and impressive gameplay (for example, bug exploitation). {\bf Disclaimer for SIGBOVIK audience: This work is 100\% real.}

On a scale from ``the title starts with Toward'' to ``Donald Knuth has finally finished the 8\th\ volume on the subject'', this work is a 3. The purpose of this paper is mainly as a careful record of the current status for repeatability and further development on this important research subject. A short video version of this paper is available for those that hate reading, at \verb+http://tom7.org/mario+, and is the more fun way to consume the results. This page also contains audiovisual material that makes this work more entertaining (for example, its output) and source code.

@@

The objective function, the algorithm to deduce it, the search strategy, and its implementation are all interesting and will be discussed in that order. To set the stage, I begin with a description of the NES hardware and emulation of it.

\subsection{The NES hardware and emulation}
The NES is based around an 8-bit processor running at 1.79~MHz, the Ricoh 2A03. 8 bits is really small. You can see them all right here: 00001111. It's no coincidence that each controller also has 8 buttons: Up, Down, Left, Right, Select, Start, B and A. It has only 2048 bytes of general purpose RAM. (There is also some special purpose RAM for graphics, which we ignore in this work.) 2048 bytes is really small. You can see them all in Figure~\ref{fig:bytes2048}. As a result, NES programs are written to use memory efficiently and straightforwardly; usually there are fixed memory locations used for all the critical game facts like the player's health, number of lives, coordinates on the screen, and so on. For example, in Super Mario Bros., the single byte at location \verb+0x757+ contains the number of lives the player has. The location \verb+0x75F+ contains the current world, and \verb+0x760+ the current level.

@@

There are a number of emulators for NES. These work by simulating the NES hardware, for example with a 2048-byte array for its memory, and simulating the steps of its 2A03 processor on some ROM, and hooking a keyboard or joystick into the 8 bits of input. (There are of course many details to work out! But in essence emulation is jut that.) This process is completely deterministic, so it is possible to record the sequence of inputs (the inputs can only be read once per video frame, so this sequence is 60 \comment{ XXX CHECK } bytes per second) and play them back and get the same result. This also means that an input sequence can be computed in non-real time, either much slower or much faster than a NES would normally run. In this work we use the FCEUX emulator, which is popular for its accuracy and advanced tools.

@@

Bytes in memory (and sometimes 16- and 32-bit words) can contain interesting game facts like the player's position in the level or score. The central idea of this paper is to use (only) the value of memory locations to deduce when the player is ``winning''. The things that a human player perceives, like the video screen and sound effects, are completely ignored. As an additional simplification, we assume that winning always consists of a value {\it going up}---either the position in the level getting larger, the score getting larger, the number of lives, the world or level number getting bigger, etc.

This is actually a little bit too naive; for example, Mario's overall progress through the game is represented by a pair. You start in World 1-1 and the underground level that comes next is World 1-2 (we'll call this $w=1$ and $\ell=2$). But after you discover the princess is in another castle in World 1-4, the next level is 2-1.\footnote{In case you never realized this, it is time to learn that the legendary ``Minus World'' of {\tt\, -1} is not actually a negative world, but World 36-1 being incorrectly rendered because there is no glyph for the 36\th\ digit. The trick used to get to the Minus World just happens to leave the value 36 in that memory location rather than initializing it to a useful value. The ROM does not contain data for world 36 so it just interprets garbage data as a description of the world.} This can't be represented as a single byte going up (sometimes the second part $\ell$ goes down when we get to a new first part $w$), but it can be represented as a lexicographic order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1, \ell_1 \rangle < \langle w_2, \ell_2$ if $w_1 = w_2$ and $\ell_1 < \ell_2$, or if $w_1 < w_2$ no matter the values of $\ell_1$ and $\ell_2$. This matches our intuitive idea and is also mathematically nice. It also generalizes multi-byte encodings of things like your score (which can be larger than 8 bits and so is often stored in 16 or 32), including both big-endian and little-endian representations.\footnote{A possible additional simplification would be to just take lexicographic orderings over bits, which then generalizes to 8-bit bytes. This is probably too crazy, but just right now I am sort of feeling like maybe I should try it, though it may be the beer.}

More importantly, it allows the combination of semantically unrelated bytes, like: $\langle$ world, level, screen inside the world, $x$ position on the screen $\rangle$ or $\langle$ world, lives, low byte of score $\rangle$. Many orderings may describe gameplay. These orderings may be temporarily violated in normal play: Although the score always goes up, Mario's $x$ position may temporarily decrease if he needs to navigate around some obstacle.\footnote{Note to self: Maybe we should give a much higher score to globally preserved objectives than to locally preserved ones. But that may presuppose that the input represents a whole playthrough?} So, to ``faithfully'' represent gameplay, we will generate a set of lexicographic orderings on memory locations, with the idea that they ``generally go up'' but not necessarily at every step. These orderings will also have weights. The next section describes how we take a sequence of player inputs and deduce the orderings.

@@

@@

On a scale from simple to fancy, this algorithm is a 3. Given those
definitions, the idea is to start with the empty prefix, which is
always a tight valid lexicographic ordering but usually not maximal.
We then pick a tight extension that is valid; if none exists then we are
done and have a maximal tight valid ordering. Now we have a new tight valid
ordering, and we repeat.

@@

\begin{figure*}[htp]
  \begin{code}
    (* Prefix is the prefix so far (int list)
       and remain is the list of memory locations
       that we can still consider. Invt is that
       prefix is a tight valid ordering on M. Returns
       the memory locations from remain that are
       a tight extension of the prefix. *)
    fun candidates (prefix, remain) =
      let lequal = (* list of indices $i$ where
                      $M_i =_{\tt prefix} M_{i+1}$ *)
      let notgreater = (* members $x$ of remain where
                          $M_i[x] > M_{i + 1}[x]$ is
                          not true for any $i$ in
                          {\tt lequal} *)
      let tight = (* members $y$ of notgreater where
                     $M_i[x] < M_{i + 1}[x]$ is true
                     for some $i$ in {\tt lequal} *)
      in tight

@@

{\bf Skipping.}\,\, To avoid being confused by RAM initialization and
menu, I ignore all memories up until the first input by the player.
Including these would be especially suspicious because the RAM's
contents are not even well-defined until the program writes something
to them.\footnote{Emulators tend to fix them to a specific pattern so
  that emulation is completely deterministic, but in real hardware
  they are truly uninitialized, or retain the values from the last
  reset or game inserted. Some semi-famous tricks involve removing
  and inserting game cartridges while the system is running in order
  to take advantage of this.}

% XXXX good place for some graphs.
{\bf Slicing.}\,\, I generate 50 orderings for $M_1 \ldots M_n$; the
whole recording starting from the first keypress. During gameplay some
values really are completely nondecreasing, like the player's score
and world/level pair. I also generate 3 orderings for each tenth of
the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and $M_{n/10+1} \ldots
M_{2n/10}$, etc. The intention is to capture orderings that are rarely
violated, or sections of the game with unusual objectives (e.g. a
minigame or swimming level). Orderings generated this way look pretty
random, and on a scale from solid to suspicious, I can't vouch for
them. Then I generate objectives from non-consecutive memories that
are evenly spread out through the observations: Ten objectives
chosen from every 100\th\ memory, starting from the 0\th\ frame,
1\st\ frame, 2\nd\ frame, and so on up to the 9\th{}. Similarly
for every 250\th\ frame, and a single objective for memory sliced
to every 1000\th\ frame, with start positions of 0--9.

{\bf Weighting.},\, To reduce the importance of randomness in the
learning process, and the arbitrariness of the slicing, each objective
function is also assigned a weight. An ideal objective function takes
on its minimal value on the first frame and maximal on the last
(according to the ordering), and increases steadily throughout the
observed memories. This is ideal because it allows us to just follow
the gradient to reach good states. A bad objective function freqently
regresses (recall that although we produce valid orderings, an
ordering that is valid on some slice may not be valid for the whole
sequence of memories). To weight an objective $L$, we first collect
all of the values (the vector of values of the memory locations $L_1
\ldots L_k$) it takes on throughout the observations. These may not
obey the ordering. We then sort the values lexicographically and
remove duplicates.\footnote{Note to self: I'm not sure how to justify
  removing duplicates here. It makes [0, 1, 1, 1, 1, 1, 10, 11] look
  the same as [0, 1, 10, 11], which is probably not desirable?} Call
this $V$. Now we can compute the {\em value fraction} for $L$ on some
$M$: Extract the vector of locations $M[L_1], M[L_2], \ldots, M[L_k]$
and find the lowest index $i$ in $V$ where the vector is less than or
equal to $V_i$. The value fraction {\tt VF} is $i/|V|$, which is the
normalized value of ``how big'' $M$ is, according to $L$, compared to
all the values we ever saw for it. The value fraction is defined and
in $[0, 1]$ for all memories in the observed set.\footnote{It is not
  defined when the memory is greater than all observed memories, which
  we will encounter later. The code returns $(i+1)/|V|$ in that case,
  which is as good as anything.} This gives us the ability to compare
objectives on an absolute scale.\footnote{This is certainly not the
  only way to do it, and it has some questionable properties like
  ignoring the magnitude of change. But it is very simple.} Weighting
an objective is now simple:
%
$$ \Sigma_{i=1}^{n-1}  {\tt VF}(M_{i+1}) - {\tt VF}(M_{i}) $$
%
We sum the differences in value functions for each consecutive pair of
memories. In the case of the ideal function this is $\Sigma_{i=1}^{n-1}
1/n$, which approaches $1$.\footnote{The code implements a different
  weighting scheme, which sums from $i=0$ and keeps track of the last
  value fraction at each step, thus treating the value fraction of the
  nonexistent memory $M_0$ as 0. This is wrong, because the first
  value fraction may be very high, which credits the objective with a
  positive value (e.g. 1) for that first part of the sum. Objectives
  that start high on the first frame are not ideal; in fact, the worst
  objectives start high on the first frame and steadily {\em
    decrease}. I plan to correct this bug and study if it has any
  noticeable consequences.} Objectives are constrained to have
non-negative weights (I set the value to 0 if negative, which
effectively disables it).

% XXX good place to give some weights for some objectives.

We save the objectives and their weights to a file and then we are
done with the easy part.

@@

\section{Now you're playing with power}

@@

It's important to know that almost all the CPU time in all the
algorithms discussed in this paper is spent emulating NES frames;
it takes about XXX ms to process a single step. Lots of engineering
effort goes into reducing the number of frames the algorithms emulate.

@@

One of the smallest NES games at a mere 24kb, Pac-Man is a classic
that I won't belabor. It has a fairly straightforward objective---your
score---and there is no timer or anything to worry about but the
ghosts. A good planner could probably solve Pac-Man with just score as
an objective (keep in mind that with RAM inspection and because of
ghost movement, this is not just a simple matter of using A$^{*}$ or
something to plan a path through the Euclidean grid---states aren't
repeated hardly ever). Automating this game works pretty well; Pac-Man
doesn't have much trouble avoiding ghosts except when they trap him.
Play is pretty strange: He doesn't necessarily grab nearby dots (hard
to explain) and often switches direction in the middle of a corridor,
unlike human play which is usually efficient sweeps of the dots when
ghosts aren't near. However, automating has a huge advantage over human
players with respect to ghosts, and Pac-Man is alarmingly fearless. He
chases ghosts down corridors, turns straight into them, knowing that
they will change direction in time to not run into him (this makes
the time travel advantage quite stark). Figure~\ref{fig:pacman} shows
gratuitous daredeviling as he ducks into a tiny sliver of space between
two ghosts, then again, and survives.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.75 \linewidth]{pacman1}
\includegraphics[width=0.75 \linewidth]{pacman2}
\includegraphics[width=0.75 \linewidth]{pacman3}
\end{center}\vspace{-0.1in}
\caption{Pac-Man showing utter disregard for the ghosts's personal
  space. This occurs around frame 6700 of the output movie. Pac-Man
  slips into the space between Blinky and Inky to touch the wall,
  comes out unharmed, then momentarily teases Clyde before escaping
  to vibrate some more in empty corridors.}
\label{fig:pacman}
\end{figure}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}
% \usepackage{inlinebib}

@@

\section{Introduction}
The Nintendo Entertainment System is probably the best video game console, citation {\it not} needed. Like many, I have spent thousands of hours of my life playing NES games, including several complete playthroughs of classics like Super Mario Bros., Bionic Commando, Bubble Bobble, and other favorites. By the year 2013, home computers have become many orders of magnitude faster and more capacious than the NES hardware. This suggested to me that it may be time to automate the playing of NES games, in order to save time.\footnote{Rather, to replace it with time spent programming.} In this paper I present a generic technique for automating the playing of NES games. The approach is practical on a single computer, and succeeds on several games, such as Super Mario Bros.. The approach is amusingly elegant and surprisingly effective, requires no detailed knowledge of the game being played, and is capable of novel and impressive gameplay (for example, bug exploitation). {\bf Disclaimer for SIGBOVIK audience: This work is 100\% real.}

On a scale from ``the title starts with Toward'' to ``Donald Knuth has finally finished the 8\th\ volume on the subject,'' this work is a 3. The purpose of this paper is mainly as a careful record of the current status for repeatability and further development on this important research subject. A short video version of this paper is available for those that hate reading, at \verb+http://tom7.org/mario+, and is the more fun way to consume the results. This page also contains audiovisual material that makes this work more entertaining (for example, its output) and source code.

@@

The objective function, the algorithm to deduce it, the search
strategy, and its implementation are all interesting and will be
discussed in that order. I then discuss the results of using the
approach to automate several NES games. To set the stage, I begin with
a description of the NES hardware and emulation of it.

\subsection{The NES hardware and emulation}
The NES is based around an 8-bit processor running at 1.79~MHz, the Ricoh 2A03. 8 bits is really small. You can see them all right here: 00001111. It's no coincidence that each controller also has 8 buttons: Up, Down, Left, Right, Select, Start, B and A. It has only 2048 bytes of general purpose RAM. (There is also some special purpose RAM for graphics, which we ignore in this work.) 2048 bytes is really small. You can see them all in Figure~\ref{fig:bytes2048}. As a result, NES programs are written to use memory efficiently and straightforwardly; usually there are fixed memory locations used for all the critical game facts like the player's health, number of lives, coordinates on the screen, and so on. For example, in Super Mario Bros., the single byte at location \verb+0x757+ contains the number of lives the player has. The location \verb+0x75F+ contains the current world, and \verb+0x760+ the current level. The NES outputs 60.0988 frames per second, which we will just call 60 in this paper.

@@

There are a number of emulators for NES. These work by simulating the
NES hardware, for example with a 2048-byte array for its memory, and
simulating the steps of its 2A03 processor on some ROM, and hooking a
keyboard or joystick into the 8 bits of input. (There are of course
many details to work out! But in essence emulation is just that.) This
process is completely deterministic, so it is possible to record the
sequence of inputs (the inputs can only be read once per video frame,
so this sequence is 60 bytes per second) and play them back and get
the same result. This also means that an input sequence can be
computed in non-real time, either much slower or much faster than a
NES would normally run. In this work we use the FCEUX\cite{FCEUX}
emulator, which is popular for its accuracy and advanced tools.

@@

Bytes in memory (and sometimes 16- and 32-bit words) can contain
interesting game facts like the player's position in the level or
score. The central idea of this paper is to use (only) the value of
memory locations to deduce when the player is ``winning''. The things
that a human player perceives, like the video screen and sound
effects, are completely ignored. As an additional simplification, we
assume that winning always consists of a value {\it going up}---either
the position in the level getting larger, the score getting larger,
the number of lives, the world or level number getting bigger, and so
on.

This is actually a little bit too naive; for example, Mario's overall
progress through the game is represented by a pair. You start in World
{\tt 1-1} and the underground level that comes next is World {\tt 1-2}
(we'll call this $w=1$ and $\ell=2$). But after you discover the
princess is in another castle in World {\tt 1-4}, the next level is
{\tt 2-1}.\footnote{In case you never realized this, it is time to learn
  that the legendary ``Minus World'' of {\tt\, -1} is not actually a
  negative world, but World {\tt 36-1} being incorrectly rendered because
  there is no glyph for the 36\th\ digit. The trick used to get to the
  Minus World just happens to leave the value 36 in that memory
  location rather than initializing it to a useful value. The ROM does
  not contain data for world 36 so it just interprets garbage data as
  a description of the world.} This can't be represented as a single
byte going up (sometimes the second part $\ell$ goes down when we get
to a new first part $w$), but it can be represented as a lexicographic
order on the pair $\langle w, \ell \rangle$; that is, $\langle w_1,
\ell_1 \rangle < \langle w_2, \ell_2 \rangle$ if $w_1 = w_2$ and $\ell_1 <
\ell_2$, or if $w_1 < w_2$ no matter the values of $\ell_1$ and
$\ell_2$. This matches our intuitive idea and is also mathematically
nice. It also generalizes multi-byte encodings of things like your
score (which can be larger than 8 bits and so is often stored in 16 or
32), including both big-endian and little-endian
representations.\footnote{A possible additional simplification would
  be to just take lexicographic orderings over bits, which then
  generalizes to 8-bit bytes. This is probably too crazy, but just
  right now I am sort of feeling like maybe I should try it, though it
  may be the beer.}

More importantly, it allows the combination of semantically unrelated
bytes, like: $\langle$\!\!\!\! world, level, screen inside the world,
$x$ position on the screen \!\!$\rangle$ or $\langle$\!\!\!\! world,
lives, low byte of score \!\!$\rangle$. Many orderings may describe
gameplay. These orderings may be temporarily violated in normal play:
Although the score always goes up, Mario's $x$ position may
temporarily decrease if he needs to navigate around some
obstacle.\footnote{Note to self: Maybe we should give a much higher
  score to globally preserved objectives than to locally preserved
  ones. But that may presuppose that the input represents a whole
  playthrough?} So, to ``faithfully'' represent gameplay, we will
generate a set of lexicographic orderings on memory locations, with
the idea that they ``generally go up'' but not necessarily at every
step. These orderings will also have weights. The next section
describes how we take a sequence of player inputs and deduce the
orderings.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-global}
\end{center}\vspace{-0.1in}
\caption[lof not used]{
  A single maximal tight valid lexicographic ordering for my
  4,000-frame input training data to Super Mario Bros.. This function
  is globally nondecreasing, and is the decimal memory locations
  $\langle$\!~232, 57, 73, 74, 75, 115, 130, 155, 32, 184, 280,
  491, 506, 1280, 1281, 1282, 1283, 1288, 1290, 1337, 1338, 1339,
  1384, 1488, 1490, 1496, 1497, 1498, 1499, 1514, 1873, 1882, 1888,
  1904, 1872, 1906, 112, 113, 114, 2009, 2010, 2011, 1539~\!$\rangle$.

  This is not a great objective function; there are long spans where
  all the memories are equal according to it, and the nice smooth slopes
  are happening during level transitions where the game is ignoring
  inputs (they are probably timers counting up each frame, which is
  why they are so smooth). Other slicing produces better objectives.

  For reasons unknown---I just discovered this while generating the
  figure---{\em all} of the objective functions learned with this
  method, regardless of the nondeterministic choices, appear to have
  this same curve, despite using different memory locations. It may be
  that they are different permutations of bytes that all change
  simultaneously, only on the frames where there are jumps in this
  picture, and there are no other orderings that are tight, valid, and
  maximal. This is still surprising and warrants investigation. }
\label{fig:global}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-tenths}
\end{center}\vspace{-0.1in}
\caption{Ten objective functions trained on different tenths of the
  4,000 inputs for Super Mario Bros.. These functions are normalized
  against the range of all values they take on during the movie; you
  can see that most are increasing locally most of the time, but many
  drop back to zero around the 3100\th\ frame, when Mario reaches
  world {\tt 1-2}. Within its 400-frame window, each objective is
  guaranteed to be nondecreasing.}
\label{fig:tenths}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-strides}
\end{center}\vspace{-0.1in}
\caption{Ten objective functions trained on every 100\th\ memory,
  starting at frame 0, frame 1, and so on up to frame 10. Normalized
  as usual. These objectives exhibit excellent global progress, but
  are locally very noisy. Among the methods I used to generate
  objectives, this one produces the best results (based on my
  intuition about what a good objective function looks like).
}
\label{fig:strides}
\end{figure}

@@

On a scale from simple to fancy, the algorithm to generate $L$ from
$M$ is a 3. Given those definitions, the idea is to start with the
empty prefix, which is always a tight valid lexicographic ordering but
usually not maximal. We then pick a tight extension that is valid; if
none exists then we are done and have a maximal tight valid ordering.
Otherwise we have a new tight valid ordering, and we repeat.

@@

\begin{figure*}[htp]
  \begin{code}
    (* Prefix is the prefix so far (int list) and remain is the list of memory
       locations that we can still consider. Invariant is that prefix is a
       tight valid ordering on $M$. Returns the memory locations from remain 
       that are a tight extension of the prefix. *)
    fun candidates (prefix, remain) =
      let lequal = (* list of indices $i$ where
                      $M_i =_{\tt\! prefix} M_{i+1}$ *)
      let notgreater = (* members $x$ of remain where
                          $M_i[x] > M_{i+1}[x]$ is
                          not true for any $i$ in
                         {\tt lequal} *)
      let tight = (* members $y$ of notgreater where
                     $M_i[x] < M_{i+1}[x]$ is true
                     for some $i$ in{\tt lequal} *)
      in tight

    \

@@

\paragraph{Skipping.} To avoid being confused by RAM initialization and
menu, I ignore all memories up until the first input by the player.
Including these would be especially suspicious because the RAM's
contents are not even well-defined until the program writes something
to them.\footnote{Emulators tend to fix them to a specific pattern so
  that emulation is completely deterministic, but in real hardware
  they are truly uninitialized, or retain the values from the last
  reset or game inserted. Some semi-famous tricks involve removing and
  inserting game cartridges while the system is running in order to
  take advantage of this.}

\paragraph{Slicing.} I generate 50 orderings for $M_1 \ldots M_n$; the
whole recording starting immediately after the first keypress. During
gameplay some values really are completely nondecreasing, like the
player's score and world/level pair. Figure~\ref{fig:global} shows
what a global ordering looks like. I also generate 3 orderings for
each tenth of the memory sequence, e.g. $M_1 \ldots M_{n/10}$ and
$M_{n/10+1} \ldots M_{2n/10}$, etc. The intention is to capture
orderings that are rarely violated, or sections of the game with
unusual objectives (e.g. a minigame or swimming level). Orderings
generated this way look pretty random, and on a scale from solid to
suspicious, I can't vouch for them. Then I generate objectives from
non-consecutive memories that are evenly spread out through the
observations: Ten objectives chosen from every 100\th\ memory,
starting from the 0\th\ frame, 1\st\ frame, 2\nd\ frame, and so on up
to the 9\th{}. Similarly for every 250\th\ frame, and a single
objective for memory sliced to every 1000\th\ frame, with start
positions of 0--9. The intention is to capture objectives that grow
slowly over the course of a playthrough, without getting distracted by
local noise.

\paragraph{Weighting.} To reduce the importance of randomness in the
learning process, and the arbitrariness of the slicing, each objective
function is also assigned a weight. An ideal objective function takes
on its minimal value on the first frame and maximal on the last
(according to the ordering), and increases steadily throughout the
observed memories. This is ideal because it allows us to just follow
the gradient to reach good states. A bad objective function freqently
regresses (recall that although we produce valid orderings, an
ordering that is valid on some slice may not be valid for the whole
sequence of memories). To weight an objective $L$, we first collect
all of the values (the vector of values of the memory locations $L_1
\ldots L_k$) it takes on throughout the observations. These may not
obey the ordering. We then sort the values lexicographically and
remove duplicates.\footnote{Note to self: I'm not sure how to justify
  removing duplicates here. It makes [0, 1, 1, 1, 1, 1, 10, 11] look
  the same as [0, 1, 10, 11], which is probably not desirable?} Call
this $V$. Now we can compute the {\em value fraction} for $L$ on some
$M$: Extract the vector of locations $M[L_1], M[L_2], \ldots, M[L_k]$
and find the lowest index $i$ in $V$ where the vector is less than or
equal to $V_i$. The value fraction {\tt VF} is $i/|V|$, which is the
normalized value of ``how big'' $M$ is, according to $L$, compared to
all the values we ever saw for it. The value fraction is defined and
in $[0, 1)$ for all memories in the observed set.\footnote{It is not
  defined when the memory is greater than all observed memories, which
  we will encounter later. The code returns $|V|/|V| = 1$ in that case,
  which is as good as anything.} This gives us the ability to compare
objectives on an absolute scale.\footnote{This is certainly not the
  only way to do it, and it has some questionable properties like
  ignoring the magnitude of change. But it is very simple.} Weighting
an objective is now simple:
%
$$ \Sigma_{i=1}^{n-1}  {\tt VF}(M_{i+1}) - {\tt VF}(M_{i}) $$
%
We sum the differences in value functions for each consecutive pair of
memories. In the case of the ideal function this is $\Sigma_{i=1}^{n-1}
1/n$, which approaches $1$. Of course, when you think about it,
this is actually the same as
%
\[
\begin{array}{c}
({\tt VF}(M_1) - {\tt VF}(M_0)) + ({\tt VF}(M_2) - {\tt VF}(M_1)) \,+ \\
\vdots \\
({\tt VF}(M_{m-1}) - {\tt VF}(M_{m-2})) + ({\tt VF}(M_{m}) - {\tt VF}(M_{m}))
\end{array}
\]
%
and all but $-{\tt VF}(M_0)$ and ${\tt VF}(M_m)$ cancel out. This
means that the weight we use is just the final value minus the initial
value, regardless of what happens in-between.\footnote{I think this
  can be improved, for example by taking the deviation from the ideal
  linear objective.} The mean value theorem or something is probably
relevant here. {\bf Lesson about being careful:} I only realized that
this code was kind of fishy when I started writing it up for SIGBOVIK.
Not only did it loop over all the deltas as in the $\Sigma$ expression
above, but it also summed from $i=0$ and kept track of the last value
fraction at each step, thus treating the value fraction of the
nonexistent memory $M_0$ as 0. This is wrong, because the first value
fraction may be very high, which credits the objective with a positive
value (e.g. 1) for that first part of the sum. Objectives that start
high on the first frame are not ideal; in fact, the worst objectives
start high on the first frame and steadily {\em decrease}. After
writing this up I corrected it to the simple expression ${\tt VF}(M_m)
- {\tt VF}(M_0)$ and the result was a huge breakthrough in the
performance of the algorithm! I had spent much time tuning the {\tt
  playfun} search procedure (Section~\ref{sec:playfun}) and not
investigated whether the objectives were being weighted properly. More
on this later, but the lesson is: Bugs matter, and thinking about your
code and explaining it is a good way to find bugs.

Objectives are constrained to have non-negative weights (I set the
value to 0 if negative, which effectively disables it). We save the
objectives and their weights to a file and then we are done with the
easy part.

@@

\section{Now you're playing with power} \label{sec:playfun}

@@

It's important to know that almost all the CPU time in all the
algorithms discussed in this paper is spent emulating NES frames; it
takes about 500\textmu s to process a single step. Lots of
engineering effort goes into reducing the number of frames the
algorithms emulate.

@@

One of the smallest NES games at a mere 24kb, Pac-Man is a classic
that I won't belabor. It has a fairly straightforward objective---your
score---and there is no timer or anything to worry about but the
ghosts. A good planner could probably solve Pac-Man with just score as
an objective (keep in mind that with RAM inspection and because of
ghost movement, this is not just a simple matter of using A$^{*}$ or
something to plan a path through the Euclidean grid---states aren't
repeated hardly ever). Automating this game works pretty well; Pac-Man
doesn't have much trouble avoiding ghosts except when they trap him.
Play is pretty strange: He doesn't necessarily grab nearby dots (hard
to explain) and often switches direction in the middle of a corridor,
unlike human play which is usually efficient sweeps of the dots when
ghosts aren't near. However, automating has a huge advantage over
human players with respect to ghosts, and Pac-Man is alarmingly
fearless. He chases ghosts down corridors, turns straight into them,
knowing that they will change direction in time to not run into him
(this makes the time travel advantage quite stark).
Figure~\ref{fig:pacman} shows gratuitous daredeviling as he ducks in
and out of a tiny sliver of space between two ghosts, then does it
again, and survives.

% ends with 13 dots remaining

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.75 \linewidth]{pacman1} \\[0.3em]
\includegraphics[width=0.75 \linewidth]{pacman2} \\[0.3em]
\includegraphics[width=0.75 \linewidth]{pacman3}
\end{center}\vspace{-0.1in}
\caption{Pac-Man showing utter disregard for the ghosts's personal
  space. This occurs around frame 6700 of the output movie. Pac-Man
  slips into the space between Blinky and Inky to touch the wall,
  comes out unharmed, then momentarily teases Clyde before escaping
  to vibrate some more in empty corridors.}
\label{fig:pacman}
\end{figure}

\subsection{Bubble Bobble}

Let's take a journey to the cave of monsters! This lovely game
features two stout dinosaurs called Bub and Bob, who jump around in a
series of single-screen caves. You can shoot bubbles to encase
monsters and then pop the bubble to turn them into fruit or other
treasure; clearing all the monsters takes you to the next stage.
Bubbles are also necessary for navigating around, since you can bounce
on your own bubbles when holding the jump button.

I was surprised that automating this game works at all, since there is
no obvious spatial gradient like in Super Mario Bros., and few local
greedy rewards like in Pac-Man. Bub initially wastes his lives (a
common theme in games where the respawn interval is low---it just
looks like a fast way of warping to the corner). When he's near
monsters he shows good tactics in shooting and popping them
simultaneously, even doing so while facing away from them, which I
didn't know was possible. At the beginning of the movie he prefers to
hide in the bottom-left corner, but soon starts jumping all around the
level, bouncing on his own bubbles and so on, almost like a real
player would. He beats XXX levels, more than I would have guessed.
Since it takes two start button presses to enter the game, the second
START is part of the input motifs. Amusingly, Bub uses this to his
advantage to change the synchronization of his futures, and when the
situation gets really tight, he'll pause the game until a good
future comes around (Figure~\ref{fig:bubblebobble}). Paused states
probably look like modest progress since memory locations like the
music cursor are still advancing.

XXX what finally happens?

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.95 \linewidth]{bubblebobble}
\end{center}\vspace{-0.1in}
\caption{Bub navigates this game surprisingly well. Once he's on his
  last life, he becomes careful and pauses the game when things are
  looking grim---here pausing for about a thousand frames, burning
  through the futures until one randomly comes along that looks good.
  He then unpauses and executes that good future, killing three of
  these monsters in short order.}
\label{fig:bubblebobble}
\end{figure}

\section{Future Work}

It is famous last words, but I actually intend to keep working on this
project. Because of SIGBOVIK crunch pressure (and discovering some
bugs as I was writing the paper) the approach got much better very
recently and I'm mostly limited by CPU hours until conference. It's
today in a state where whenever I run it on a new game I see something
delightful. Even just running it on more of the NES classics is
worthwhile. However, I have lots of ideas about how to extend the
technique, either to make it more beautiful or to make it work better:

\paragraph{Parameter reduction.} I've already pointed out the places
where there are mysterious constants in the code. Although the fact
that the same constants work on many different games without tuning
is some solace, it would really be nicer to remove these. One of the
hardest to remove is the length of the futures explored. XXX thoughts
on how to improve this?

\paragraph{Multiple players, multiple games.} Other than the particulars
of the way it's built (\verb+vector<uint8>+ everywhere), there's no
reason why the technique is limited to a single player's input. In a
game like Bubble Bobble or Contra the two players can collaborate in
interesting ways, and planning both simultaneously would probably lead
to occasional awesome results. For example, in Contra, it might be
that one player is shooting enemies for the other player, the bullets
arriving from across the screen just in time to save him as he
blithely jumps into danger. Another clever feat from the Tool Assisted
Speedrun community is a sequence of inputs that beats multiple
different {\em games} simultaneously. For example, human geniuses used
tools to beat Mega Man (Mega Men?) 3, 4, 5, and 6 all at the same time
using the exact same input sequence sent to all four
games.\cite{BaxterTAS} I think the algorithms in this paper apply
directly---it's just a matter of multiplexing the inputs to each of
the games, and then taking the appropriate operation (min, max, sum) of
their objective functions to produce a final objective function. The
main obstacle is the architecture of the underlying emulator, which
can only load one game into its global variables at once.

\bibliographystyle{plain}
% \bibliographycomment{} % nothing
\bibliography{paper}
